---
title: "Homework 4"
output: html_notebook
author: Hannah Marr
---

II. R

6. Create a boxplot of the distribution of Plain pizza prices over the years using the cleaned dataset pizza data from class. Then, create a histogram displaying the frequency of Plain’ pizza prices in 2022. Provide the R code. (2 points)

```{r}
# Load the necessary libraries for data manipulation and visualization
library(tidyverse)  # Collection of R packages for data science
library(tidyr)      # Specifically for tidying data
library(dplyr)      # For data manipulation (select, filter, group, etc.)
```
```{r}
# Read the pizza data from a CSV file into a dataframe
pizza_raw_data <- read.csv("/Users/hannahmarr/Desktop/Tufts/DATA200/Labs/Pizza_NYC.csv")

# Select specific columns from the dataset for analysis (Name, location, date, price, and style)
pizza_data <- pizza_raw_data %>%
  select(Name, location_lat, location_lng, Date, Year, Price, Style)

# Remove rows with missing values (NAs) to clean the data
pizza_data <- drop_na(pizza_data)

# Check for any remaining missing values in each column after cleaning
colSums(is.na(pizza_data))

# Get the dimensions (number of rows and columns) of the cleaned dataset
dim(pizza_data)
```
```{r}
# Filter the pizza_data dataframe into a new dataframe with only Plain pizza
style_plain <- pizza_data %>% 
  filter(Style == 'Plain')
head(style_plain)
```
```{r}
# Create a boxplot to visualize price of plain pizza by year
boxplot(style_plain$Price ~ style_plain$Year,
        main = "Plain Pizza Price by Year",
        xlab = "Year",
        ylab = "Price",
        col = "tomato4",
        las = 1)
```
```{r}
# Subset the data for the year 2022
plain_pizza_2022 <- subset(style_plain, Year == 2022)  
head(plain_pizza_2022)
```
```{r}
# Plot a histogram of the Price column for the year 2022
hist(plain_pizza_2022$Price,
      main = "Histogram of Plain Pizza Prices in 2022",
      xlab = "Price",
      col = "coral1",  # Fill color for the bars
      border = "black")  # Border color for the bars
```

7. Create a subset named pizza pepperoni that only contains Pepperoni pizzas. Group the data by year and calculate the maximum, minimum, and average prices. Then, use a line chart to visualize these trends. You can either create three separate lines for maximum, minimum, and average prices, or combine them into a single chart that includes all three lines. Provide the R code (3 points)

```{r}
# Subset the data for the style pepperoni
pizza_pepperoni <- subset(pizza_data, Style == 'Pepperoni')  
head(pizza_pepperoni)
```

```{r}
# Group by year and calculate max, min, and average prices
pepperoni_stats <- pizza_pepperoni %>%
  group_by(Year) %>%
  summarise(
    max_price = max(Price, na.rm = TRUE),
    min_price = min(Price, na.rm = TRUE),
    avg_price = mean(Price, na.rm = TRUE)
  )

#View the summarized data
head(pepperoni_stats)
```
```{r}
# Plot the trends
plot(pepperoni_stats$Year, pepperoni_stats$max_price, type = 'l', col = 'red',
     ylim = c(min(pepperoni_stats$min_price), max(pepperoni_stats$max_price)),
     xlab = 'Year', ylab = 'Price', main = 'Pepperoni Pizza Price Trends')
lines(pepperoni_stats$Year, pepperoni_stats$min_price, type = 'l', col = 'blue')
lines(pepperoni_stats$Year, pepperoni_stats$avg_price, type = 'l', col = 'green')

# Add a legend
legend('topright', legend = c('Max Price', 'Min Price', 'Avg Price'),
       col = c('red', 'blue', 'green'), lty = 1)
```

